Verified Nanopasses for Compiling Conditionals

### 2025-06-18 paper.tex
tweak

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
We present a proof of correctness in Agda for four nanopasses that
translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an intermediate language
that is roughly x86 assembly with variables (prior to register
allocation).  The most interesting of these four nanopasses is a
translation of conditional expressions into goto-style control flow
that uses the continuation-oriented approach of Olivier Danvy's
one-pass transformation into monadic normal form (2003).
\end{abstract}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to a sophisticated
algorithm of \citet{Aho:1986qf} (presented in the form of an attribute
grammar) that translates Boolean expressions into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes \citep{PEYTONJONES19983}.

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that no longer contains
Boolean expressions or nested conditionals, and where jumps are
expressed using zero-arity function calls. One of the key parts of
that translation is in the compilation of a conditional expression
$(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled using a
special function $\mathcal{B}$ that takes two more arguments, the
generated code for the two branches. Thus, $\mathcal{B}$ can analyze
the expression $b$ to decide how to generate code for the
$\mathsf{if}$, or whether it is even necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa}, in which he taught the techniques that
he employed in building Chez Scheme~\citep{Dybvig:2006aa}, including
destintation-driven code generation~\cite{Dybvig:1990aa}. The idea
there was to use an extra parameter in the compilation function to
indicate how the context would use the result value of an
expression. The context might (1) ignore the result (and only care
about the expression's side effects), (2) branch based on interpreting
its result as a boolean, or (3) actually use the result
value. Relating this to Olivier's transformation functions, the
$\mathcal{E}$ function is for contexts that need the actual value and
the the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of function is isomorphic to a
function on a disjoint union.
\[
(A \to B) \times (C \to B) \cong (A + C) \to B
\]

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
We present a proof of correctness in Agda for four nanopasses that
translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The most interesting of these four nanopasses is a
translation of conditional expressions into goto-style control flow
that uses the continuation-oriented approach of Olivier Danvy's
one-pass transformation into monadic normal form (2003).
\end{abstract}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes \citep{PEYTONJONES19983}.

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that no longer contains
Boolean expressions or nested conditionals, and where jumps are
expressed using zero-arity function calls. One of the key parts of
that translation is in the compilation of a conditional expression
such as $(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled
using a special function $\mathcal{B}$ that takes two more arguments,
the generated code for the two branches. Thus, $\mathcal{B}$ can
analyze the expression $b$ to decide how to generate code for the
$\mathsf{if}$, or whether an $\mathsf{if}$ is even necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destintation-driven code
generation~\cite{Dybvig:1990aa}, which employs an extra parameter on
the compilation function that indicates how the context will use the
result value of an expression. The context might (1) ignore the result
(and only care about the expression's side effects), (2) branch based
on interpreting its result as a boolean, or (3) actually use the
result value. These options are encoded as a disjoint union.  Relating
this to Olivier's transformation functions, the $\mathcal{E}$ function
is for contexts that need the actual value and the $\mathcal{B}$
function is for contexts that branch on the expression's result. Of
course, a pair of functions is isomorphic to a function on a disjoint
union.
\[
(A \to C) \times (B \to C) \cong (A + B) \to C
\]

The CompCert compiler \citep{Leroy:2006fe}, a compiler for C verified
in Coq, uses the one-function per kind of context approach, in the
RTLgen pass that translates from the CminorSel intermediate language
to its register transfer language.
